A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2025-07-09 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: expanding with examples

\textbf{Quasi-probability representations.} Another way to represent quantum ensembles is through the use of quasiprobability distributions. While the Wigner function is probably the most widely known, there are in fact several different approach. The Wigner function represent quantum states, both pure and mixed, as a density function over position and momentum. This function integrates to one over the full space, like a probability density, but can have small negative regions, unlike a probability density. Recall that an affine combination is a linear combination where all coefficients sum to one but are not necessarily non-negative. Therefore, in general, quasi-probability distributions can be understood as affine combinations.

For the Wigner function, the expectation of polynomials of position and momentum over phase space recovers expectations of polynomials of position and momentum in symmetric order (?). (Definitions and features?) In some problems, especially in quantum optics, normal ordering and anti-normal ordering are more useful. The Glauber P representation and the Husimi Q representation work like the Wigner function except that the expectations of polynomials recover the respective order. In other circumstances, quasi-probability distributions over observables different from position and momentum are more useful. For example, the Kirkwood-Dirac is used to represent two-time correlation functions over observables defined at different times. Quasi-probability representations are also used over discrete elements that may not even correspond to eigenstates of a particular operator. For example, the SIC-POVM represents density matrices as affine combinations of pure states that are ``equally distant'' from each other, which means that no two elements are orthogonal to each other.

It should be clear that there are many ways to represent quantum states via quasi-probability distributions, and that the choice of representation is a matter of convenience for the problem at hand. All representations, though, have the following features in common. First, we have a map $W : \Ens \to Q(Z,\mathbb{R})$ that given each mixed state $\rho \in \Ens$ returns the corresponding quasi-probability function $w(z) \in Q(Z,\mathbb{R})$, where $Q(Z,\mathbb{R})$ is the function space used by the specific representation. The representation preserves convex combinations: $W(\sum_i p_i \rho_i) = \sum_i p_i W(\rho_i)$. That is, a statistical mixture of quasi-probability distributions represents the statistical mixture of the corresponding states. Since the map is linear, and expectation values are linear, one can express expectation values as integrals or sums over the quasi-probability $w(z)$.

@@

\textbf{Not about negative probability.} It is very common to conceive the main difference between classical probability and these quasi-probabilities as the inclusion of negative probability. Sometimes these negative probabilities are justified as arising from constraints on probability distributions defined on different contexts. This is a mischaracterization. For example, the Husimi Q representation provides us with a non-negative quasi-probability function. That is, the Husimi Q is a non-negative density distribution that integrates to one, yet it is not a classical probability distribution. Similarly, since we can embed the Bloch ball inside a tetrahedron, we can represent all states of a two-state quantum system by a convex combination of the vertices, leading to a non-negative quasi-probability distribution over four elements.

Conversely, we can use signed measure to represent classical distributions. Suppose we have a two state classical discrete space, which means the space of all possible ensembles is spanned by $[p, (1-p)]$. Alternatively, we can write it as $p[1,0] + (1-p)[0,1]$ which highlights that we are writing each possible probability distribution in terms of the two possible cases. But we could also express each probability distribution in terms of other two distributions, as in $\left(2p - \frac{1}{2}\right)\left[\frac{3}{4}, \frac{1}{4}\right] + \left(\frac{3}{2} - 2p\right)\left[\frac{1}{4}, \frac{3}{4}\right]$. The coefficients now span between $\left[-\frac{1}{2}, \frac{3}{2}\right]$, can be negative, but still sum to one. That is, in terms of this new affine basis, we need to use a signed probability.

Since we can represent classical states with signed probability measures, and quantum states with non-negative probability measure, the difference between classical probability and quantum quasi-probability does not lie in the use of negative probability. Instead, the crucial difference is that in classical probability all points of the sample space, and therefore disjoint regions, represent mutually exclusive events. That is, they can be prepared and observed independently from the others; we can assign some probability to some region without having to assign it to another disjoint region.\footnote{The only constraint is that the probability over the whole space is one.} This is not the case for quasi-probability in quantum mechanics. Husimi Q represents a state as decomposed over coherent states, which cannot represent mutually exclusive as they are not orthogonal. The inner product between these states will provide a constrain. In the non-negative probability representation of the Bloch sphere, none of the four corners represent valid states, as they lie outside the Bloch ball. So we cannot assign probability to a vertex without assigning some probability to another vertex.

>>>>

\textbf{Quasi-probability representations.} Another way to represent quantum ensembles is through the use of quasiprobability distributions, of which the Wigner function is probably the most widely known. The Wigner transform $W : \Ens \to L^1(\mathbb{R}^2, \mathbb{R}) $, defined as [CITATION NEEDED]
\begin{equation}
	W(\rho) = W_{\hat{\rho}} (q, p) = 2 \int_{-\infty}^{\infty} \langle x + y | \rho | x - y \rangle e^{-2 i p y / \hbar} \; \mathrm dy,
\end{equation}
takes a mixed state $\rho$ and returns a function of phase space that integrates to one, like a probability density, but can have small negative regions. Recall that an affine combination is a linear combination where all coefficients sum to one but are not necessarily non-negative. Therefore, in general, quasi-probability distributions can be understood as affine combinations. Remarkably, expectation values of polynomial of position and momentum in symmetric ordering, coincide with expectations over the Wigner function. For example
\begin{equation}
	\tr \left( \frac{1}{2}[XP+PX] \rho \right) = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} xp \; W_{\rho} (q, p) dq dp \; .
\end{equation}
Moreover, the transform is an affine map: $W(p \rho_1 + \bar{p} \rho_2 ) = p W(\rho_1) + \bar{p} W(\rho_2)$.

In some problems, especially in quantum optics, normal ordering and anti-normal ordering are more useful. The Glauber P representation and the Husimi Q representation work like the Wigner function except that the expectations of polynomials recover the respective order. In other circumstances, quasi-probability distributions over observables different from position and momentum are more useful. For example, the Kirkwood-Dirac is used to represent two-time correlation functions over observables defined at different times. Quasi-probability representations are also used over discrete elements that may not even correspond to eigenstates of a particular operator. For example, the SIC-POVM represents quantum states as affine combinations of pure states $P_i$ that are ``equally distant'' from each other, which means that no two elements are orthogonal to each other. Each ensemble $\rho$ can then be represented as
\begin{equation}
	\rho = \sum_{n=1}^{d^2} \lambda_i P_i = \sum_{n=1}^{d^2} \left[ (d^2 + 1) p_i - \frac{1}{d} \right] P_i \; ,
\end{equation}
where $p_i = \tr \left(P_i \rho\right)$. [CITATION NEEDED]

It should be clear that there are many ways to represent quantum states via quasi-probability distributions, and that the choice of representation is a matter of convenience for the problem at hand. All representations, though, have the following features in common. First, we have a map $W : \Ens \to Q(Z,\mathbb{R})$ that given each mixed state $\rho \in \Ens$ returns the corresponding quasi-probability function $w(z) \in Q(Z,\mathbb{R})$, where $Q(Z,\mathbb{R}) \subset L^1(Z,\mathbb{R})$ is the function space used by the specific representation. The space $Z$, in fact, can change from representation to representation: it is $\mathbb{R}^{2n}$ for a Wigner function, $\mathbb{C}^n$ for the Husimi Q and $[1, d^2] \subset \mathbb{N}$ for the SIC-POVM. All representations preserves convex combinations: $W(\sum_i p_i \rho_i) = \sum_i p_i W(\rho_i)$. That is, a statistical mixture of quasi-probability distributions represents the statistical mixture of the corresponding states. Since the map is linear, and expectation values are linear, one can express expectation values as integrals or sums over the quasi-probability $w(z)$.

@@

\textbf{Not about negative probability.} The inclusion of negative probability is often advertised as the main difference between classical probability and quasi-probability. [CITATION NEEDED] Sometimes these negative probabilities are justified as arising from constraints on probability distributions defined on different contexts. This is a mischaracterization. For example, the Husimi Q representation provides us with a non-negative quasi-probability function. That is, the Husimi Q is a non-negative density distribution that integrates to one, yet it is not a classical probability distribution.\footnote{Since the Husimi Q is defined as $Q_{\rho} ( \alpha ) = \frac{1}{\pi} \langle \alpha | \rho | \alpha \rangle$, it is proportional to the probability of measuring a coherent (i.e. Gaussian state) $| \alpha \rangle$, which is non-negative. } Similarly, a SIC-POVM characterizes the Bloch ball by four pure states that form a tetrahedron inscribed in the ball. Since we can also inscribe the Bloch ball inside a tetrahedron, we can represent all states of a two-state quantum system by a convex combination of the vertices, leading to a non-negative quasi-probability distribution over four elements.

Conversely, we can use signed measures to represent classical distributions. Suppose we have a two state classical discrete space, which means the space of all possible ensembles is spanned by $[p, (1-p)]$. Alternatively, we can write it as $p[1,0] + (1-p)[0,1]$ which highlights that we are writing each possible probability distribution in terms of the two possible cases. But we could also express each probability distribution in terms of other two distributions, as in $\left(2p - \frac{1}{2}\right)\left[\frac{3}{4}, \frac{1}{4}\right] + \left(\frac{3}{2} - 2p\right)\left[\frac{1}{4}, \frac{3}{4}\right]$. The coefficients now span between $\left[-\frac{1}{2}, \frac{3}{2}\right]$, can be negative, but still sum to one. That is, in terms of this new affine basis, we need to use a signed probability.

Since we can represent classical states with signed probability measures, and quantum states with non-negative probability measure, the difference between classical probability and quantum quasi-probability does not lie in the use of negative probability. Instead, the crucial difference is that in classical probability all points of the sample space, and therefore disjoint regions, represent mutually exclusive events, and all mutually exclusive events are represented by disjoint regions. Mutually exclusive events and be prepared and observed independently from the others (i.e. they are orthogonal in the sense that they maximize entropy increase during mixture); we can assign some probability to some region without having to assign it to another disjoint region.\footnote{The only constraint is that the probability over the whole space is one.} This is not the case for quasi-probability in quantum mechanics.

For example, the eigenstates $| m \rangle$ of the harmonic oscillator are orthogonal, mutually exclusive events. Their corresponding Wigner functions are
\begin{equation}
	W_m (q, p) = \frac{(-1)^m}{\pi} e^{-(q^2+p^2)} L_m (2(q^2 + p^2)).
\end{equation}
Their support is the whole $\mathbb{R}^2$ since the Laguerre polynomials $L_m$ have finitely many zeros, and therefore are not disjoint. Conversely, in a SIC-POVM we take $d^2$ pure states $P_i$ in a $d$ dimensional space, which means they cannot be all orthogonal to each other. For a Bloch sphere, the four corners of the tetrahedron are not all opposite to each other. Therefore, disjoint sets of $P_i$ do not correspond to mutually exclusive events.
